---
title: "K-nearest neighbour"
author: "Diana Escamilla"
date: "March 16, 2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##HOMEWORK 4####
#PROBLEM 1: k-nearest neighbours [100pts]
(a) We are going to implement k-nearest neighbours. Recall how this works: you're given some training data with labels. Given a new test datapoint, you decide its label by performing a majority vote among its k-nearest neighbours. Write down a skeleton or outline of how you might implement this (in R or in English/pseudocode). This needn't bear any similarity to your final program, but you should have an idea of all the pieces you'll need and how they fit together

```{r K-Nearest Neighbor Pseudocode}
#Let (Xi, Ci), where i= 1,2.....,n be data points
#Xi are the values of each data point
#Ci are the lables for Xi for each i 
#Assuming the number of classes as 'C'
#Ci is {1,2,3,..,C}  for all values of i
# Let x be a point for which label is not known and we want to find the label class using k-nearest neighbor algorithms
#steps:
#1. Calculate the Euclidean distance between the points. d(x, xi) wiht i= 1,2,...,n where d denotes the euclidean distance between the points
#2. Arrange calculated n Euclidean distances in non-decreasing order
#3. Let K be a positive integer, take the first K distances from this sorted list
#4. Find those K-points corresponding to these K-distances
#5. Let Ki denotes the number of points belonging to the ith class among K points  i.e. K =>0
#6. If ki >kj for all i different than j then put x in class i.
```
b) Install package RnavGraphImageData and use the data() function to load the digits dataset: data(digits). This dataframe contains the so-called USPS dataset, with each column having length 256, and representing an 16 × 16 image of a handwritten digit. See ?digits details. (WARNING: for some reason,digits 6 and 7 are overwritten by 5, don't let that confuse you later)
```{r Loading Dataset}
#install.packages("RnavGraphImageData")
Func1<- function(){
  library(RnavGraphImageData)
  data(digits)
  return(digits)
    }
Digits<- Func1()
```
c) It's helpful to visualize the digits, and we will write a function plot_digit to do this. It should take a vector (a column of digits) as input, and plot it using the image() function. For this you'll have to convert it to a 16 × 16 matrix (experiment with the byrow option of matrix). You can also play with the color argument of image, I used col = gray(0:255/255)
```{r Plodigit Function}
plot_digit<- function(vector){
  a<-matrix(vector, ncol = 16, nrow = 16, byrow = TRUE)
  a<- apply(a, 1, rev)
  a<- t(a)
  library(RColorBrewer)
  image(a, col= brewer.pal(n=3, name = "Greens"))
  }

resutls2<- plot_digit(Digits$V7701) ##I run one column to see if the function works

```
d) What are the index ranges of each digit in digits? Plot the first instance of each digit (since image doesn't involve ggplot you don't have to organize it in a panel). However, instead of calling your function 10 times manually, write a for loop (or use *pply). Careful: the order of digits in digits is 1,2,.,9,0 rather than 0,1,2,.. You should notice that digits 6 and 7 are just 5
```{r plot first instance of each digit}
##there are 1100 repetitions of each number
func2<- function(file){
  seq<- c(1, 1101, 2201,3301, 4401,5508,6618,7701,8801,9901) ##when running the function it prints the 10 numbers however when knitting it omits two of the 5s (6 and 7), I have to print a different number for 6 and 7 digit to be able to see them in the final file instead of the first image of the digits.
  apply(file[,seq],2,plot_digit)
  }

results3<- func2(Digits)

```

e) Write a function get_digits to convert digits to a smaller, more convenient dataset. get_digits should take two arguments: the first is a vector select_digs and the second is size. The function should return the first size elements of each digit in select_digs. Thus if get_digits is c(0,5,4) and size is 50, then the function should return a data frame of length 150, consisting of the first 50 images of 0, 5 and 4. Create a dataset my_train consisting of 100 instances of 0 and 8.
```{r get_digits}
get_digits<- function(select_digs, size){
  cols<- numeric()
  select_digs[select_digs== 0]<-10
  Firtsimage<- (select_digs-1)*1100 +1
  for (i in 1:length(Firtsimage)){
    cols<- c(cols, seq(Firtsimage[i], length.out = size))
  }
  return(Digits[,cols])
}
select_digs<- c(0,8)
size<- 100
my_train<- get_digits(select_digs, size)
```

(f) Write a function euc_dist that accepts two vectors (of any length), and returns the Euclidean distance between them. You can normalize this by the vector length if you want

```{r euclidian distances}
library(dplyr)
euc_dist<-function(a, b) (a-b)^2 %>% sum %>% sqrt
```
  
(g) Given any new digit, we want to calculate the distance to every element of my_train. Use laply from package plyr. laply accepts a list as input and returns an array obtained by appling some function fun to each element of that list. Recall that my_train is a dataframe which is just a list. If the function needs more than one argument you can pass those to laply, see the documentation and recall how '...'works in functions. You overall syntax with look like my_arr <- laply(my_list, fun, second_arg) where you plug in appropriate variable names
```{r distance between new digit and every element in my train}

func3<- function(){
  library(plyr)
  new_digit<- Digits[763]
  my_arr<- laply(my_train, euc_dist, new_digit)
  return(my_arr)
}

results3<- func3()
head(results3)
length(results3)

```

(h) Now write a function get_knn that takes 3 inputs: a number k, the training data, and a single testvector. It should return the indices of the k nearest neighbours of the test-vector in the training data. It does this in three steps: first calculate the distance to each element in the training set, and then sort them using the sort command. Calling sort with the option index.return set to TRUE also returns
the indices of the sorted elements. Finally, return the first k indices. 
```{r function get_Knn}
get_knn<- function(k, my_train, test_vector){
  distances<- laply(my_train,euc_dist,test_vector)
  indices<- sort(distances, index.return= TRUE)$ix ##allows to return the indices (positions) of the sorted elements
  return(indices[1:k])
}

```
i) Now write a function to get the majority label of the returned k indices. For this, it is useful to complement my_train from step (e) with a vector my_labels containing the corresponding labels.
```{r labels}
my_labels<- rep(c(0,8), each= 100)
get_labels<- function(kindices){
  labels<- my_labels[kindices]
  lab<- table(labels)
  return(names(which.max(lab)))
}


```
j) Wrap the previous few functions into a function my_knn that takes four inputs: k, my_train, my_labels and test_ip, and returns the predicted label of test_ip from applying k-nearest neighbours. 

```{r my_knn function}
my_knn<- function(k, my_train, my_labels, test_ip ){
  kindices<- get_knn(k, my_train, test_ip)
  label<- get_labels(kindices)
  return(label)
}
```

(k) Set k = 5. Apply your function setting test_vec to each element of my_train. How many 0's does it get wrong? How many 8's? 
```{r using my train as test_vec}

Prediction<- function(test_ip, my_train, k, my_labels) {
  predict<- laply(test_ip, my_knn, my_train = my_train, k=k, my_labels=my_labels)
  a<- paste("my_knn got wrong", sum(predict[1:100] != my_labels[1:100]), "0's."); print(a)
  b<-paste("my_knn got wrong", sum(predict[101:200] != my_labels[101:200]), "8's."); print(b)
}

Prediction(my_train, my_train, 5, my_labels)
```
(l) Apply your function to a hundred 0's and 8's from digits NOT present in my_train. How many of these does it get wrong?
```{r apply the function to digits not present in my train}
test1<- function (my_train, k, my_labels){
  test_ip<- Digits[, c(10000:10099, 7900:7999)]
  test_lables<- rep(c(0,8), each= 100)
  predict<- laply(test_ip, my_knn, my_train = my_train, k=k, my_labels=my_labels)
  a<- paste("my_knn got wrong", sum(predict[1:100] != test_lables[1:100]), "0's."); print(a)
  b<-paste("my_knn got wrong", sum(predict[101:200] != test_lables[101:200]), "8's."); print(b)
}
test1(my_train, 5, my_labels)

```
 Apply it to one hundred 5's. How many of these does it classify as 0, and how many as 8? 
```{r apply to a hundred 5s}

test2<- function (my_train, k, my_labels){
  test_ip<- Digits[, 4800:4899]
  predict<- laply(test_ip, my_knn, my_train = my_train, k=k, my_labels=my_labels)
  table(predict)
}
test2(my_train, 5, my_labels)
```
